---
title: "Lab 1 Group 8"
output:
  pdf_document:
    latex_engine: lualatex
date: "2025-01-25"
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Optimization of a two-dimensional function

Consider the two-dimensional function:

$$f(x, y) = \sin(x + y) + (x - y)^2 - 1.5x + 2.5y + 1$$

for $x \in [-1.5, 4]$ and $y \in [-3, 4]$.The aim is to search for the position $(x^*, y^*)$ of the global minimum of $f$ in the given range for $x$ and $y$.

### a) Make a contour plot for this function.
```{r echo=FALSE}

f <- function(x, y) {
  sin(x + y) + (x - y)^2 - 1.5 * x + 2.5 * y + 1
}
x_values <- seq(-1.5, 4, length.out = 200)
y_values <- seq(-3, 4, length.out = 200)
coordinates <- expand.grid(x_values, y_values)

results <- matrix(apply(coordinates, 1, function(data) f(data[1], data[2])),nrow=length(x_values))

contour(x_values, y_values, results,nlevels = 40,xlab = "x", ylab = "y", main = "Contour Plot of f(x, y)",
        col = "blue", lty = 1, lwd = 2)
```



### b) Derive the gradient and Hessian matrix for the function and write R-code for them.
```{r echo=FALSE}

gradient <- function (x, y) {
  return(matrix(c(
    cos(x + y) + 2 * x - 2 * y - 1.5, cos(x + y) + 2 * y - 2 * x + 2.5
  ), nrow = 2))
  
}

Hessian <- function (x , y) {
  return(matrix(c(
    -sin(x + y) + 2, -sin(x + y) - 2, -sin(x + y) - 2, -sin(x + y) + 2
  ), nrow = 2))
}
```

### c) Write your own function applying the Newton algorithm that has the starting value $(x_0, y_0)$ as a parameter
```{r echo=FALSE}
#stopping criteria
convergence_crit<- function(res) {
  xt1<- res[,dim(res)[2]]
  xt<- res[,dim(res)[2]-1]
  return(as.vector(t (xt1-xt) %*% (xt1-xt)))
}

Newton_algorithm <- function (x_0, y_0) {
  res <- matrix ( c(x_0,y_0), nrow=2)
   res <- cbind(res, Hessian (x_0, y_0) %*% gradient (x_0 ,y_0)) 
   
   while (convergence_crit(res) >0.001) {
     xt1<- res[,dim(res)[2]]
     res_add <- xt1 - solve (Hessian (xt1[1], xt1[2])) %*% gradient (xt1[1] ,xt1[2])
     res <- cbind(res, res_add)
   }
   return (res)
} 


```


### d) Test several starting values such that you have examples which give at least three different results.
```{r echo=FALSE}
values<- matrix (c(2,4,
                   -1,0,
                   -2,-3,
                   2,0
                   #1,2,
                   #4,0.1
                  ),
                 ncol = 2,byrow = TRUE)
for (i in 1:nrow(values)) {
  res_df<-Newton_algorithm(values[i,1],values[i,2])
  #browser()
  cat('Newton_algorithm gives a mimimum value at (',print(res_df[1,dim(res_df)[2]]), ',', print(res_df[2,dim(res_df)[2]]),'at the ',dim(Newton_algorithm(2,1))[2]-1, 'iteration for a=',values[i,1], 'b=', values[i,2],'\n')
}
```
 
### e) Investigate the candidate results which you have got. Is one of the candidates a global minimum? Which type of solutions are the other candidates?
The last solution gives (-1.59, -2.59) is a global minimum. (1.54, -0.55) is the saddle point. (2.59, 1.59) is a local maximum.

