---
title: "Lab 1 Group 8"
output:
  pdf_document:
    latex_engine: lualatex
date: "2025-01-25"
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Optimization of a two-dimensional function

Consider the two-dimensional function:

$$f(x, y) = \sin(x + y) + (x - y)^2 - 1.5x + 2.5y + 1$$

for $x \in [-1.5, 4]$ and $y \in [-3, 4]$.The aim is to search for the position $(x^*, y^*)$ of the global minimum of $f$ in the given range for $x$ and $y$.

### a) Make a contour plot for this function.
```{r echo=FALSE}

f <- function(x, y) {
  sin(x + y) + (x - y)^2 - 1.5 * x + 2.5 * y + 1
}
x_values <- seq(-1.5, 4, length.out = 200)
y_values <- seq(-3, 4, length.out = 200)
coordinates <- expand.grid(x_values, y_values)

results <- matrix(apply(coordinates, 1, function(data) f(data[1], data[2])),nrow=length(x_values))

contour(x_values, y_values, results,nlevels = 10,xlab = "x", ylab = "y", main = "Contour Plot of f(x, y)",
        col = "blue", lty = 1, lwd = 2)
```



### b) Derive the gradient and Hessian matrix for the function and write R-code for them.


### c) Write your own function applying the Newton algorithm that has the starting value $(x_0, y_0)$ as a parameter

### d) Test several starting values such that you have examples which give at least three different results.
 
### e) Investigate the candidate results which you have got. Is one of the candidates a global minimum? Which type of solutions are the other candidates?
