<<<<<<< HEAD
=======

>>>>>>> 400fcf8e92ef15b07c749749641be09e8b77afc9
---
title: "Lab 4 Group 8"
output:
  pdf_document:
    latex_engine: lualatex
<<<<<<< HEAD
date: "2025-02-18"
=======
date: "2025-02-10"
>>>>>>> 400fcf8e92ef15b07c749749641be09e8b77afc9
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<<<<<<< HEAD
## Question 1: Consider a random variable \(X\) with the following probability density function:
\[ f(x) = 120x^5 e^{-x}, \quad x > 0 \]

### Part a
Use the Metropolis–Hastings algorithm to generate 10,000 samples \(X_t\), \(t = 1, \ldots, 10,000\), from this distribution. Use a normal distribution with mean \(X_t\) and standard deviation 0.1 as the proposal distribution when you generate \(X_{t+1}\); take some starting point. Plot the chain you obtained with iterations on the horizontal axis.  Plot a histogram of the sample.
```{r echo=TRUE}
prop_dis <- function (xt) {
  return(rnorm(1, xt, 0.1))
}

f <- function(x)
{
  return(120 * x ^ 5 * exp(-x))
}


set.seed(123)
x      <- 1
chain  <- c(x)
acceptance_count <- 0

# generate first five values (change 5 to higher number for longer chain)
for (i in 1:10000)
{
  # generate first one sample from uniform distribution on circle with specific radius by rejection sampling:
  
  xcand <- prop_dis(x)        # candidate point
  # dnorm as the density of it
  R     <- f(xcand) * dnorm(x, mean = xcand, sd = 0.1) / 
    (f(x) * dnorm(xcand, mean = x, sd = 0.1))  # MH ratio
  
  ap    <- runif(1)
  if (ap < R) {
    x <- xcand
      acceptance_count <- acceptance_count + 1

  }
    
  chain <- rbind(chain, x)
}
head(chain,100)  # each row is one two-dimensional observation

plot(chain, type = "l", main = "Metropolis-Hastings Chain", xlab = "Iteration", ylab = "X")
# Plot histogram of the samples



```
What can you guess about the convergence of the chain? If there is a burn-in period, what can be the size of this period? What is the acceptance rate?
```{r}

# Plot the histogram of the samples
hist(chain, breaks = 50, main = "Histogram of Samples with Target Density", xlab = "X", probability = TRUE)

# Add the target density function
x_values <- seq(0, 15, by = 0.1)
y_values <- f(x_values)
plot(x_values, y_values, col = "blue", lwd = 2)


# Calculate acceptance rate
acceptance_rate <- acceptance_count / 10000
cat("Acceptance Rate:", acceptance_rate, "\n")

```
It seems that the convergence is not reached for this chain. The plot of X-values vs iteration number still has a great fluctuation and the plotted histogram does not have a good fitting with the true plot of target function. The X-values of histogram of sample lies in [0,10] with the highest density at around 4. While the true function plot gives v-values in [0,15] and it peaks around 5.  
By looking into the first 100 samples, we found that X-value start to stabilize around 2 from the 40th sampling. Therefore, 40 can be the size of burn-in period if there is a burn-in period.
The acceptance rate is calculated to be 0.9856.

### Part b
Perform Part a by using the chi-square distribution \(\chi^2(\lfloor X_t + 1 \rfloor)\) as a proposal distribution, where \(\lfloor x \rfloor\) is the floor function, meaning the integer part of \(x\) for positive \(x\), i.e. \(\lfloor 2.95 \rfloor = 2\).
```{r}

prop_dis2<- function (x,xt) {
  dchisq(x,df=floor(xt+1))
}

set.seed(123)
x      <- 1
chain2  <- c(x)
acceptance_count <- 0
for (i in 1:10000)
{
  # generate first one sample from uniform distribution on circle with specific radius by rejection sampling:
  
  xcand <- rchisq(1,floor(x+1))        # candidate point
  # dnorm as the density of it
  R     <- f(xcand) * prop_dis2(x,xcand)/ 
    (f(x) * prop_dis2(xcand,x))  # MH ratio
  
  ap    <- runif(1)
  if (ap < R) {
    x <- xcand
      acceptance_count <- acceptance_count + 1

  }
    chain2 <- rbind(chain2, x)

  }
head(chain2)  # each row is one two-dimensional observation

plot(chain2, type = "l", main = "Metropolis-Hastings Chain", xlab = "Iteration", ylab = "X")

# Plot the histogram of the samples
hist(chain2, breaks = 50, main = "Histogram of Samples with Target Density", xlab = "X", probability = TRUE)
plot(x_values, y_values, col = "blue", lwd = 2)
# Calculate acceptance rate
acceptance_rate <- acceptance_count / 10000
cat("Acceptance Rate:", acceptance_rate, "\n")
```


### Part c
Suggest another proposal distribution (can be a normal or chi-square distribution with other parameters or another distribution) with the potential to generate a good sample. Perform Part a with this distribution.
A normal distribution with standard deviation 5 is suggested. 
```{r echo=TRUE}
test_deviation = 5
prop_dis3 <- function (xt) {
  return(rnorm(1, xt, test_deviation))
}

f <- function(x)
{
  return(120 * x ^ 5 * exp(-x))
}


set.seed(123)
x      <- 1
chain3  <- c(x)
acceptance_count <- 0

# generate first five values (change 5 to higher number for longer chain)
for (i in 1:10000)
{
  # generate first one sample from uniform distribution on circle with specific radius by rejection sampling:
  
  xcand <- prop_dis3(x)        # candidate point
  # dnorm as the density of it
  R     <- f(xcand) * dnorm(x, mean = xcand, sd = test_deviation) / 
    (f(x) * dnorm(xcand, mean = x, sd = test_deviation))  # MH ratio
  
  ap    <- runif(1)
  if (ap < R) {
    x <- xcand
      acceptance_count <- acceptance_count + 1

  }
    
  chain3 <- rbind(chain, x)
}
head(chain3,10)  # each row is one two-dimensional observation

plot(chain3, type = "l", main = "Metropolis-Hastings Chain", xlab = "Iteration", ylab = "X")
# Plot histogram of the samples

# Plot the histogram of the samples
hist(chain3, breaks = 50, main = "Histogram of Samples with Target Density", xlab = "X", probability = TRUE)

# Add the target density function
x_values <- seq(0, 15, by = 0.1)
y_values <- f(x_values)
plot(x_values, y_values, col = "blue", lwd = 2)


# Calculate acceptance rate
acceptance_rate <- acceptance_count / 10000
cat("Acceptance Rate:", acceptance_rate, "\n")


```
This chain is generally considered to reach convergence for the stable plot and its good fitting of the target distribution. X -values reaches 2 and begin to stabilize at the 5th iteration, indicting a very short burn- in period. Calculated acceptance rate is 0.4737.
### Part d
Compare the results of Parts a, b, and c and make conclusions.
Chain 1 has not yet reached convergence with unstable plot and bad fitting of the true plot. Chain 2 and 3 are considered to reach convergence and they both have good representation of the distribution. While chain 2 has a higher acceptance rate 0.6062 comparing to 0.4737 in chain 3. As is suggest by Hoeting, Chapter 7.3, the preferable acceptance rate is 44%. Therefore, chain 3 is the best out of all 3 chains.

### Part e
Estimate
\[ E(X) = \int_0^\infty x f(x) \, dx \]
using the samples from Parts a, b, and c.
```{r}
estimate_expectation <- function(chain) {
  product <- chain * sapply(chain, f)
  
  E_X_estimate <- mean(product)
  return(E_X_estimate)
}
chian1_exp <- estimate_expectation(chain)
chian2_exp <- estimate_expectation(chain2)
chian3_exp <- estimate_expectation(chain3)

cat("E(X) for Chain 1: ", chian1_exp, "\n")
cat("E(X) for Chain 2: ", chian2_exp, "\n")
cat("E(X) for Chain 3: ", chian3_exp, "\n")
```
The expectation values are 9529.828, 9812.96 and 9529.865 for chain 1-3 respectively. 

### Part f
The distribution generated is in fact a gamma distribution. Search in the literature/internet and define the actual value of the integral. Compare it with the one you obtained.

For a Gamma distribution with shape parameter \( \alpha \) and scale parameter \( \beta \), the following properties hold:

1. **Mean**:
\[
E(X) = \alpha \beta
\]

2. **Variance**:
\[
\text{Var}(X) = \alpha \beta^2
\]

- **Shape parameter \( \alpha \)**:
\[
\alpha = \frac{E(X)^2}{\text{Var}(X)}
\]

- **Scale parameter \( \beta \)**:
\[
\beta = \frac{\text{Var}(X)}{E(X)}
\]
```{r}
chian1_var <-var(chain)
chian2_var <- var(chain2)
chian3_var <- var(chain3)


# Compute alpha and beta for each chain
beta1 <- chian1_exp / chian1_var
alpha1 <- chian1_exp * beta1

beta2 <- chian2_exp / chian2_var
alpha2 <- chian2_exp * beta2

beta3 <- chian3_exp / chian3_var
alpha3 <- chian3_exp * beta3

# Print results
cat("Chain 1: alpha =", alpha1, ", beta =", beta1, "\n")
cat("Chain 2: alpha =", alpha2, ", beta =", beta2, "\n")
cat("Chain 3: alpha =", alpha3, ", beta =", beta3, "\n")
```

## Question 2: Gibbs Sampling

Let \(X = (X_1, X_2)\) be a bivariate distribution with density \(f(x_1, x_2) \propto 1\{x_1^2 + wx_1x_2 + x_2^2 < 1\}\) for some specific \(w\) with \(|w| < 2\). \(X\) has a uniform distribution on some two-dimensional region. We consider here the case \(w = 1.999\) (in Lecture 4, the case \(w = 1.8\) was shown).
=======

a)
>>>>>>> 400fcf8e92ef15b07c749749641be09e8b77afc9
```{r}
w  <- 1.999
xv <- seq(-1, 1, by=0.01) * 1/sqrt(1-w^2/4)  # a range of x1-values, where the term below the root is non-negative (compare Lecture 4)
plot(xv, xv, type="n", xlab=expression(x[1]), ylab=expression(x[2]), las=1)
# ellipse
lines(xv, -(w/2)*xv-sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
lines(xv, -(w/2)*xv+sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
<<<<<<< HEAD
```


### Part a
Draw the boundaries of the region where \(X\) has a uniform distribution. You can use the code provided on the course homepage and adjust it.

### Part b
What is the conditional distribution of \(X_1\) given \(X_2\) and that of \(X_2\) given \(X_1\)?

### Part c
Write your own code for Gibbs sampling the distribution. Run it to generate \(n = 1000\) random vectors and plot them into the picture from Part a. Determine \(P(X_1 > 0)\) based on the sample and repeat this a few times (you need not plot the repetitions). What should be the true result for this probability?

### Part d
Discuss why the Gibbs sampling for this situation seems to be less successful for \(w = 1.999\) compared to the case \(w = 1.8\) from the lecture.

### Part e
We might transform the variable \(X\) and generate \(U = (U_1, U_2) = (X_1 - X_2, X_1 + X_2)\) instead. In this case, the density of the transformed variable \(U = (U_1, U_2)\) is again a uniform distribution on a transformed region (no proof necessary for this claim). Determine the boundaries of the transformed region where \(U\) has a uniform distribution on. You can use that the transformation corresponds to \(X_1 = \frac{U_2 + U_1}{2}\) and \(X_2 = \frac{U_2 - U_1}{2}\) and set this into the boundaries in terms of \(X_i\). Plot the boundaries for \((U_1, U_2)\). Generate \(n = 1000\) random vectors with Gibbs sampling for \(U\) and plot them. Determine \(P(X_1 > 0) = P\left(\frac{U_2 + U_1}{2} > 0\right)\). Compare the results with Part c.
=======


```


b) X1 | X2 :

$$
x_1 \mid x_2=\frac{-1.999 x_2 \pm \sqrt{\left(1.999 x_2\right)^2-4\left(x_2^2-1\right)}}{2}
$$


X2 | X1:
$$
x_2 \mid x_1=\frac{-1.999 x_1 \pm \sqrt{\left(1.999 x_1\right)^2-4\left(x_1^2-1\right)}}{2}
$$
c)
```{r}
get_limits <- function(fixed_val, w, var) {
  a <- 1
  b <- w * fixed_val
  c_val <- fixed_val^2 - 1
  discriminant <- b^2 - 4 * a * c_val
  if (discriminant < 0) return(c(NA, NA))
  sqrt_d <- sqrt(discriminant)
  lower <- (-b - sqrt_d) / (2 * a)
  upper <- (-b + sqrt_d) / (2 * a)
  return(c(lower, upper))

}

gibbs_sample <- function(n, w) {
  samples <- matrix(0, nrow = n, ncol = 2)
  x1 <- 0
  x2 <- 0
  for (i in 1:n) {
    #browser()
    # fix X2 and update X1 | X2
    limits_x1 <- get_limits(x2, w, "x1")
    if (any(is.na(limits_x1))) break
    #sample between the lower and upper limits
    x1 <- runif(1, limits_x1[1], limits_x1[2])
    # fix X2 and update X2 | X1
    limits_x2 <- get_limits(x1, w, "x2")
    if (any(is.na(limits_x2))) break
    x2 <- runif(1, limits_x2[1], limits_x2[2])
    samples[i, ] <- c(x1, x2)
  }
  return(samples)
}
set.seed(123)
samples <- gibbs_sample(1000, 1.999)
plot(samples, pch = 20, col = rgb(0, 0, 1, 0.3), main = "Gibbs Samples")

prob <- mean(samples[,1] > 0)
cat("Estimated P(X1 > 0):", prob)

```

d) from the plot above, we can see that x1 and x2 is highly correlated, The ellipse becomes extremely narrow compared to W = 1.8 and approaches a line. so the sample value could only in a very narrow range.




e) Through variable conversion, the value range of X1 and X is broadened, the sampling efficiency is improved, and the estimation is closer to the true value，But because of w equals 1.999 which results in a slight asymmetry,so the actual value may be closer to 0.5
```{r}


gibbs_sample_transformed = function(n,w) {
  samples_u = matrix(0, nrow = n, ncol = 2)
  u1 <- 0
  u2 <- 0
  for (i in 1:n) {
    # fix U2 and update U1 | U2
    u2_current <- u2
    u1_min <- -sqrt((4 - 3.999 * u2_current^2) / 0.001)
    u1_max <- sqrt((4 - 3.999 * u2_current^2) / 0.001)
    u1 <- runif(1, u1_min, u1_max)
    # fix U1 and update U2 | U1
    u1_current <- u1
    u2_min <- -sqrt((4 - 0.001 * u1_current^2) / 3.999)
    u2_max <- sqrt((4 - 0.001 * u1_current^2) / 3.999)
    u2 <- runif(1, u2_min, u2_max)
    samples_u[i, ] <- c(u1, u2)
  }
  return(samples_u)
}

samples_u <- gibbs_sample_transformed(1000, 1.999)
#browser()
samples_x_transformed <- cbind(
  (samples_u[,1] + samples_u[,2]) / 2,
  (samples_u[,2] - samples_u[,1]) / 2
)
plot(samples_x_transformed, pch = 20, col = rgb(0, 0, 1, 0.3), main = "Gibbs Samples")

prob_u <- mean(samples_x_transformed[,1] > 0)
cat("Transformed P(X1 > 0):", prob_u) 
  
```


>>>>>>> 400fcf8e92ef15b07c749749641be09e8b77afc9
