
---
title: "Lab 4 Group 8"
output:
  pdf_document:
    latex_engine: lualatex
date: "2025-02-10"
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

a)
```{r}
w  <- 1.999
xv <- seq(-1, 1, by=0.01) * 1/sqrt(1-w^2/4)  # a range of x1-values, where the term below the root is non-negative (compare Lecture 4)
plot(xv, xv, type="n", xlab=expression(x[1]), ylab=expression(x[2]), las=1)
# ellipse
lines(xv, -(w/2)*xv-sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
lines(xv, -(w/2)*xv+sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)


```


b) X1 | X2 :

$$
x_1 \mid x_2=\frac{-1.999 x_2 \pm \sqrt{\left(1.999 x_2\right)^2-4\left(x_2^2-1\right)}}{2}
$$


X2 | X1:
$$
x_2 \mid x_1=\frac{-1.999 x_1 \pm \sqrt{\left(1.999 x_1\right)^2-4\left(x_1^2-1\right)}}{2}
$$
c)
```{r}
get_limits <- function(fixed_val, w, var) {
  a <- 1
  b <- w * fixed_val
  c_val <- fixed_val^2 - 1
  discriminant <- b^2 - 4 * a * c_val
  if (discriminant < 0) return(c(NA, NA))
  sqrt_d <- sqrt(discriminant)
  lower <- (-b - sqrt_d) / (2 * a)
  upper <- (-b + sqrt_d) / (2 * a)
  return(c(lower, upper))

}

gibbs_sample <- function(n, w) {
  samples <- matrix(0, nrow = n, ncol = 2)
  x1 <- 0
  x2 <- 0
  for (i in 1:n) {
    #browser()
    # fix X2 and update X1 | X2
    limits_x1 <- get_limits(x2, w, "x1")
    if (any(is.na(limits_x1))) break
    #sample between the lower and upper limits
    x1 <- runif(1, limits_x1[1], limits_x1[2])
    # fix X2 and update X2 | X1
    limits_x2 <- get_limits(x1, w, "x2")
    if (any(is.na(limits_x2))) break
    x2 <- runif(1, limits_x2[1], limits_x2[2])
    samples[i, ] <- c(x1, x2)
  }
  return(samples)
}
set.seed(123)
samples <- gibbs_sample(1000, 1.999)
plot(samples, pch = 20, col = rgb(0, 0, 1, 0.3), main = "Gibbs Samples")

prob <- mean(samples[,1] > 0)
cat("Estimated P(X1 > 0):", prob)

```

d) from the plot above, we can see that x1 and x2 is highly correlated, The ellipse becomes extremely narrow compared to W = 1.8 and approaches a line. so the sample value could only in a very narrow range.




e) Through variable conversion, the value range of X1 and X is broadened, the sampling efficiency is improved, and the estimation is closer to the true value，But because of w equals 1.999 which results in a slight asymmetry,so the actual value may be closer to 0.5
```{r}


gibbs_sample_transformed = function(n,w) {
  samples_u = matrix(0, nrow = n, ncol = 2)
  u1 <- 0
  u2 <- 0
  for (i in 1:n) {
    # fix U2 and update U1 | U2
    u2_current <- u2
    u1_min <- -sqrt((4 - 3.999 * u2_current^2) / 0.001)
    u1_max <- sqrt((4 - 3.999 * u2_current^2) / 0.001)
    u1 <- runif(1, u1_min, u1_max)
    # fix U1 and update U2 | U1
    u1_current <- u1
    u2_min <- -sqrt((4 - 0.001 * u1_current^2) / 3.999)
    u2_max <- sqrt((4 - 0.001 * u1_current^2) / 3.999)
    u2 <- runif(1, u2_min, u2_max)
    samples_u[i, ] <- c(u1, u2)
  }
  return(samples_u)
}

samples_u <- gibbs_sample_transformed(1000, 1.999)
#browser()
samples_x_transformed <- cbind(
  (samples_u[,1] + samples_u[,2]) / 2,
  (samples_u[,2] - samples_u[,1]) / 2
)
plot(samples_x_transformed, pch = 20, col = rgb(0, 0, 1, 0.3), main = "Gibbs Samples")

prob_u <- mean(samples_x_transformed[,1] > 0)
cat("Transformed P(X1 > 0):", prob_u) 
  
```


