---
title: "Bayesian Learning - Computer Lab 1"
output:
  pdf_document:
    latex_engine: lualatex
date: "2025-04-14"
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## 1. Daniel Bernolli

Let $y_1, \ldots, y_n \mid \theta \sim \text{Bern}(\theta)$, and assume that you have obtained a sample with $f = 35$ failures in $n = 78$ trials. Assume a $\text{Beta}(\alpha_0, \beta_0)$ prior for $\theta$ and let $\alpha_0 = \beta_0 = 7$.

### (a)

Draw 10000 random values (nDraws = 10000) from the posterior $\theta \mid y \sim \text{Beta}(\alpha_0 + s, \beta_0 + f)$, where $y = (y_1, \ldots, y_n)$, and verify graphically that the posterior mean $\mathbb{E}[\theta \mid y]$ and standard deviation $\text{SD}[\theta \mid y]$ converges to the true values as the number of random draws grows large.

Give model is a bernoulli trial and prior has a $\text{Beta}(\alpha_0, \beta_0)$ distribution, posterior $ \sim \text{Beta}(\alpha_0 + s, \beta_0 + f)$, where s=(n-f)=53 and f=35. Thus: 
    posterior $ \sim \text{Beta}(\alpha_0 + s, \beta_0 + f)$
    posterior $ \sim \text{Beta}(50, 42)$


```{r eval=TRUE}
nr <- 10000
set.seed(123456)
res <- rbeta(nr,50,42)
me <- numeric(nr)
std <- numeric(nr)

for (i in 1: nr) {
  me[i] <- mean(res[1:i])
  std[i] <- sqrt(var (res [1:i]))

}

plot(me, type = 'l')
abline(h=50/(50+42), ,col ='red')

plot(std, type = 'l')
abline(h=sqrt(50*42/(50+42)^2/(42+50+1)),col ='red')

```

### (b) 
Draw 10000 random values from the posterior to compute the posterior probability  
\( \Pr(\theta > 0.5 \mid y) \) and compare with the exact value from the Beta posterior. [Hint: use `pbeta()` to calculate the exact value].

```{r eval=TRUE}
sorted <- res [res>0.5]
# probability from sampling  
pr <- length(sorted) / nr  

# true probability  
tr_pr <- 1- pbeta (0.5, 50, 42)
```
Posterior probability from sampling: 
```{r eval=TRUE}
print(pr)
```

Exact value from the Beta posterior: 
```{r eval=TRUE}
print(tr_pr)

```
The sampling gives very similar values comparing with the exact value from Beta posterior. 

### (c) 
Draw 10000 random values from the posterior of the odds \( \phi = \frac{\theta}{1 - \theta} \) by using the previous random draws from the Beta posterior for \( \theta \) and plot the posterior distribution of \( \phi \). [Hint: `hist()` and `density()` can be utilized].

```{r eval=TRUE}
odds <- res / (1-res) 
hist(odds, breaks = 30, probability = TRUE, main = "Posterior of Odds", xlab = "Odds")
lines(density(odds), col = "blue", lwd = 2)
```


