---
title: "Bayesian Learning - Computer Lab 2"
output:
  pdf_document:
    latex_engine: lualatex
date: "2025-04-28"
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## 1. Linear and polynomial regression

The dataset `temp_linkoping.csv` contains daily average temperatures (in degree Celsius) in Linköping between July 2023 and June 2024. Import the dataset in R.\
The response variable is `temp` and the covariate `time`, which is defined as

$$
\text{time} = \frac{\text{the number of days since the beginning of the observation period}}{365}.
$$

A Bayesian analysis of the following quadratic regression model is to be performed:

$$
\text{temp} = \beta_0 + \beta_1 \cdot \text{time} + \beta_2 \cdot \text{time}^2 + \varepsilon, \quad \varepsilon \overset{\text{iid}}{\sim} \mathcal{N}(0, \sigma^2).
$$\

#### (a)

Use the conjugate prior for the linear regression model. The prior hyperparameters $\mu_0$, $\Omega_0$, $\nu_0$ and $\sigma_0^2$ shall be set to sensible values. Start with

$$
\mu_0 = (0, 100, -100)^T, \quad \Omega_0 = 0.01 \cdot I_3, \quad \nu_0 = 1, \quad \sigma_0^2 = 1.
$$

Check if this prior agrees with your prior opinions by simulating draws from the joint prior of all parameters and for every draw compute the regression curve.\
This gives a collection of regression curves; one for each draw from the prior.\
Does the collection of curves look reasonable? If not, change the prior hyperparameters until the collection of prior regression curves agrees with your prior beliefs about the regression curve.\
*Hint: R package `mvtnorm` can be used and your Scaled-Inv-*$\chi^2$ simulator of random draws from Lab 1.

```{r eval=TRUE}
data <- read.csv('temp_linkoping.csv')
library (mvtnorm)

# define inverse chi square function
mu_0 <- c(15, -50, 50)
nu_0<- 1
sig_0_sq <- 2
omega_0 <- diag(3)*0.8
 rinvchisq <- function(n, df, scale) {
  df * scale / rchisq(n, df)
  }
```

Join prior for bete given

```{r eval=TRUE}
# sort our time and time^2
data['time^2'] <- data[3]^2
data['time^0'] <- data[3]^0
time <- as.matrix(data[, c(6,3,5)])
temp <- as.matrix(data[, 4])

curve <-  matrix(NA, nrow = 366, ncol = 100)

for  (i in 1: 100) {
  sig_sq_sp<- rinvchisq(1,nu_0, sig_0_sq)
  samp <- rmvnorm(n=1, mean= mu_0, sigma = sig_sq_sp*solve(omega_0))
  pred  <- time %*% as.numeric(samp)
  curve [,i] <- pred
}

plot(time[,2], temp, pch = 16, col = 'gray', xlab = "Time", ylab = "Temperature")
for (i in 2:100) {
  lines(data[,3], curve[,i], col = rgb(0,0,1,alpha=0.2))  
}
```

#### (b)

Write a function that simulates draws from the joint posterior distribution of $\beta_0$, $\beta_1$, $\beta_2$ and $\sigma^2$.

i.  Plot a histogram for each marginal posterior of the parameters.

ii. Make a scatter plot of the temperature data and overlay a curve for the posterior median of the regression function

$$
f(\text{time}) = \mathbb{E}[\text{temp}|\text{time}] = \beta_0 + \beta_1 \cdot \text{time} + \beta_2 \cdot \text{time}^2,
$$

i.e., the median of $f(\text{time})$ is computed for every value of `time`.\
In addition, overlay curves for the 90% equal tail posterior probability intervals of $f(\text{time})$, i.e., the 5 and 95 posterior percentiles of $f(\text{time})$ are computed for every value of `time`.\
Does the posterior probability interval contain most of the data points? Should they?

##### Part 1

Since we use a Normal-Inverse-Chi-Squared prior on $(\beta, \sigma^2)$, where:

-   $\beta \mid \sigma^2 \sim \mathcal{N}(\mu_0, \sigma^2 \Omega_0^{-1})$
-   $\sigma^2 \sim \text{Inv-}\chi^2(\nu_0, s_0)$

Given observed data $(X, y)$, the joint posterior distribution of $\beta$ and $\sigma^2$ is: - The marginal posterior of $\sigma^2$ is:

$$
\sigma^2 \mid y \sim \text{Inv-}\chi^2(\nu_n, \sigma_n)
$$

with: $$
\nu_n = \nu_0 + n
$$

and $$
\nu_n \sigma_n^2 = \nu_0 \sigma_0^2 + y^\top y + \mu_0^\top \Omega_0 \mu_0 - \mu_n^\top \Omega_n \mu_n
$$ where the updated parameters are:

$$
\Omega_n = \Omega_0 + X^\top X
$$

$$
\mu_n = \Omega_n^{-1}(\Omega_0 \mu_0 + X^\top y)
$$

-   The conditional posterior of $\beta$ given $\sigma^2$ and data is:

$$
\beta \mid \sigma^2, y \sim \mathcal{N}(\mu_n, \sigma^2 \Omega_n^{-1})
$$

```{r eval=TRUE}
set.seed(123456)
sample_posterior <- function (number ,nu_0, sig_0_sq, omega_0) {
  # The marginal posterior of sigma

nu_n <- nu_0+366
omega_n <- t(time) %*%time+omega_0
mu_n <- solve(omega_n) %*% (omega_0%*%mu_0 +t(time)%*%(temp))
sig_n_sq <- as.numeric(1/nu_n*(nu_0*sig_0_sq+
                                 t(temp)%*%temp+
                                 t(mu_0)%*%omega_n%*%mu_0-
                                 t(mu_n)%*%omega_n%*%mu_n))

# sampled results
# joint posterior of sigma square given data
  sig2_marg<- rinvchisq(number,nu_n, sig_n_sq)
  
# joint posterior of beta given sigma square and data   
beta_marg <- matrix(nrow=number, ncol=3)
colnames(beta_marg) = c('beta0', 'beta1','beta2')

 for (i in 1: number) {
     beta_marg[i,] <- as.vector(rmvnorm(n=1, mean= mu_n, sigma = sig2_marg[i]*solve(omega_n)))
 }
return(list(sig2_marg = sig2_marg, beta_marg = beta_marg))
}

res <-sample_posterior(number=10000 ,nu_0=nu_0, sig_0_sq=sig_0_sq, omega_0=omega_0)
sig2_marg <- res$sig2_marg
beta_marg <- res$beta_marg

```

Then we can plot the histogram for $\sigma^2$ given data:

```{r eval=TRUE}
hist(sig2_marg, breaks = 20)
```

The histogram for $\beta_0$ given $\sigma^2$ and data:

```{r eval=TRUE}
hist(beta_marg[,1], breaks = 20)
```

The histogram for $\beta_1$ given $\sigma^2$ and data:

```{r eval=TRUE}
hist(beta_marg[,2], breaks = 20)
```

The histogram for $\beta_2$ given $\sigma^2$ and data:

```{r eval=TRUE}
hist(beta_marg[,1], breaks = 20)

```

##### Part 2

```{r eval=TRUE}
pred_temp <- time%*%t(beta_marg)
pred_temp_meadian <- apply(pred_temp,1, quantile, probs = c(0.05,0.95) )

plot(time[,2], temp, pch = 16, col = 'gray', xlab = "Time", ylab = "Temperature",
     main = "Posterior Median and 90% Prediction Interval")
points(time[,2], pred_temp_meadian[1,], col ='red')
points(time[,2], pred_temp_meadian[2,], col ='blue')

```

#### (c)

It is of interest to locate the time with the lowest expected temperature (i.e., the time where $f(\text{time})$ is minimal). Let's call this value $\tilde{x}$.\
Use the simulated draws in (b) to simulate from the posterior distribution of $\tilde{x}$.\
You can solve this analytically or numerically.\
*Hint: The regression curve is a quadratic polynomial. Given each posterior draw of* $\beta_0$, $\beta_1$, and $\beta_2$, you can find a simple formula for $\tilde{x}$.

To find the time point $\tilde{x}$ at which the function reaches its minimum , we compute the derivative with respect to time:

$$
\frac{d}{d\,\text{time}} f(\text{time}) = \beta_1 + 2\beta_2 \cdot \text{time}
$$

Setting the derivative equal to zero:

$$
\beta_1 + 2\beta_2 \cdot \tilde{x} = 0
$$

Solving for $\tilde{x}$ gives:

$$
\tilde{x} = -\frac{\beta_1}{2\beta_2}
$$

```{r eval=TRUE}
xbar <- -beta_marg[,2]/2/beta_marg[,3]
hist(xbar, breaks = 20)
```

#### (d)

Say now that you want to estimate a polynomial regression of order 10, but you suspect that higher order terms may not be needed, and you worry about overfitting the data.\
Suggest a suitable prior that mitigates this potential problem and motivate your choice.\
Repeat task (a) for the selected prior and the higher order polynomial to see if it behaves as expected.\
*Hint: The task is to specify* $\mu_0$ and $\Omega_0$ in a suitable way.

To reflect our prior belief that lower-order terms are likely important, we assign relatively low precision values (small diagonal values (1) in Ω₀) to them.On the other hand, for higher-order terms (e.g., degree 5 to 10), we set much higher prior precision (1000), encouraging these coefficients to be small unless the data provides strong evidence otherwise.

It is shown that the curve fit the actual data better than what we have in task (a), but no significant improvement is obtained than what we have from task (b) where we have even lower order. Therefore, we believe that higher order terms are not needed.    

```{r eval=TRUE}
# define inverse chi square function
mu_0_new <- c(25, -60, 0.01,-1,60,10,0,0,0,0,0)
nu_0<- 1
sig_0_sq <- 2
omega_0_new <- diag(c(rep(1, 6), rep(1000, 5))) 

#construct poly terms
x <- data[, 3]
poly_terms <- sapply(0:10, function(p) x^p)
colnames(poly_terms) <- paste0("time^", 0:10)
time_new <- as.matrix(poly_terms)

# sort our time and time^2

curve_new <-  matrix(NA, nrow = 366, ncol = 100)

for  (i in 1: 100) {
  sig_sq_sp<- rinvchisq(1,nu_0, sig_0_sq)
  samp_new <- rmvnorm(n=1, mean= mu_0_new, sigma = sig_sq_sp*solve(omega_0_new))
  pred_new  <- time_new %*% as.numeric(samp_new)
  curve_new [,i] <- pred_new
}

plot(time[,2], temp, pch = 16, col = 'gray', xlab = "Time", ylab = "Temperature"#,ylim=range(curve_new)
     )
for (i in 2:100) {
  lines(data[,3], curve_new[,i], col = rgb(0,0,1,alpha=0.2)) 
}
```
