---
title: "Bayesian Learning - Computer Lab 3"
output:
  pdf_document:
    latex_engine: lualatex
date: "2025-0-13"
author:
  - Liuxi Mei
  - Xiaochen Liu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## 1. Gibbs sampling for the logistic regression

Consider again the logistic regression model in problem 2 from the previous computer lab 2. Use the prior $\beta \sim \mathcal{N}(0, \tau^2 I)$, where $\tau = 3$.

(a) Implement (code!) a Gibbs sampler that simulates from the joint posterior $p(\omega, \beta|x)$ by augmenting the data with Polya-gamma latent variables $\omega_i, i = 1, \dots, n$. The full conditional posteriors are given on the slides from Lecture 7. Evaluate the convergence of the Gibbs sampler by calculating the Inefficiency Factors (IFs) and by plotting the trajectories of the sampled Markov chains.

(b) Repeat the same task as in (a), but now only use the first m observations of the dataset. Do this for all $m \in \{10, 40, 80\}$.
```{r eval=TRUE}

```


## 2. Metropolis Random Walk for Poisson regression

Consider the following Poisson regression model:

\[
y_i | \beta \sim \text{Poisson}\left( \exp(x_i^T \beta) \right), \quad i = 1, \dots, n,
\]

where $y_i$ is the count for the $i$th observation in the sample and $x_i$ is the $p$-dimensional vector with covariate observations for the $i$th observation. 

Use the dataset `eBayNumberOfBidderData_2025.dat`. This dataset contains observations from 700 eBay auctions of coins. The response variable is `nBids` and records the number of bids in each auction. The remaining variables are features/covariates ($x$):


## 2. Metropolis Random Walk for Poisson regression

Consider the following Poisson regression model:

\[
y_i | \beta \sim \text{Poisson}\left( \exp(x_i^T \beta) \right), \quad i = 1, \dots, n,
\]

where $y_i$ is the count for the $i$th observation in the sample and $x_i$ is the $p$-dimensional vector with covariate observations for the $i$th observation.

The dataset contains observations from 700 eBay auctions of coins. The response variable is `nBids` and records the number of bids in each auction. The remaining variables are features/covariates ($x$):

- **Const**: Intercept
- **PowerSeller**: Equal to 1 if the seller is selling large volumes on eBay
- **VerifyID**: Equal to 1 if the seller is a verified seller by eBay
- **Sealed**: Equal to 1 if the coin was sold in an unopened envelope
- **MinBlem**: Equal to 1 if the coin has a minor defect
- **MajBlem**: Equal to 1 if the coin has a major defect
- **LargNeg**: Equal to 1 if the seller received a lot of negative feedback from customers
- **LogBook**: Logarithm of the book value of the auctioned coin according to expert sellers (Standardized)
- **MinBidShare**: Ratio of the minimum selling price (starting price) to the book value (Standardized).

### (a) Maximum Likelihood Estimator of β in Poisson Regression

Use `glm()` to obtain the maximum likelihood estimator of $\beta$ for the eBay dataset. **Hint:** Remember that `glm()` automatically includes the intercept, so don’t input the covariate `Const`. After fitting the model, identify which covariates are significant.

### (b) Bayesian Analysis of Poisson Regression

Assume the prior:

\[
\beta \sim N(0, 100 \cdot (X^T X)^{-1}),
\]

where $X$ is the $n \times p$ covariate matrix. This is Zellner's $g$-prior. Assume the posterior density is approximately multivariate normal:

\[
\beta | y \sim N(\tilde{\beta}, J_y^{-1}(\tilde{\beta})),
\]

where $\tilde{\beta}$ is the posterior mode and $J_y(\tilde{\beta})$ is the negative Hessian at the posterior mode. Use numerical optimization (via `optim()`) to compute $\tilde{\beta}$ and $J_y^{-1}(\tilde{\beta})$. 

### (c) Simulate from the Posterior of β using Metropolis Algorithm

Now simulate from the actual posterior of $\beta$ using the Metropolis algorithm. Program a general function for the Metropolis algorithm that generates random draws from any arbitrary posterior density. The proposal density will be the multivariate normal density:

\[
\theta_p | \theta_{i-1} \sim N(\theta_{i-1}, c \cdot \Sigma),
\]

where $\Sigma = J_y^{-1}(\tilde{\beta})$ is obtained in (b), and $c$ is a tuning parameter. The function should allow users to supply their own posterior density function, not necessarily for Poisson regression, making it a general-purpose Metropolis function.

After writing the Metropolis function, use it to sample from the posterior of $\beta$ for the Poisson regression in the eBay dataset. Assess MCMC convergence by graphical methods.

## (d) Simulate from the Predictive Distribution using MCMC Draws

Use the MCMC draws from part (c) to simulate from the predictive distribution of the number of bidders in a new auction with the following characteristics:

- PowerSeller = 1
- VerifyID = 0
- Sealed = 1
- MinBlem = 0
- MajBlem = 1
- LargNeg = 0
- LogBook = 1.3
- MinBidShare = 0.7

Simulate the number of bidders and plot the predictive distribution. What is the probability of having no bidders in this new auction? 

### 3. Time Series Models in Stan

#### (a) Simulate Data from the AR(1)-process

Write an R function that simulates data from the AR(1)-process:

\[
x_t = \mu + \phi \cdot (x_{t-1} - \mu) + \epsilon_t, \quad \epsilon_t \sim N(0, \sigma^2),
\]

for given values of $\mu$, $\phi$, and $\sigma^2$. Start the process at $x_1 = \mu$ and then simulate values for $x_t$ for $t = 2, 3, \dots, T$, and return the vector $x_{1:T}$ containing all time points. Use $\mu = 5$, $\sigma^2 = 9$, and $T = 300$. Simulate several realizations of $x_{1:T}$ for values of $\phi$ between $-1$ and $1$, the stationary interval for the AR(1)-process. Include a plot of at least one realization of $x_{1:T}$ in the report. 

What effect does the value of $\phi$ have on $x_{1:T}$?

#### (b) Simulate Two AR(1)-Processes and Perform Bayesian Inference in Stan

Use your function from part (a) to simulate two AR(1)-processes:

- $x_{1:T}$ with $\phi = 0.4$
- $y_{1:T}$ with $\phi = 0.98$

Now, treat the simulated vectors as synthetic data and treat the values of $\mu$, $\phi$, and $\sigma^2$ as unknown parameters. Implement Stan code that samples from the posterior of the three parameters, using suitable non-informative priors of your choice. 

- **i.** Report the posterior mean, 95% credible intervals, and the number of effective posterior samples for the three inferred parameters for each of the simulated AR(1)-processes. Are you able to estimate the true values?

- **ii.** For each of the two datasets, evaluate the convergence of the samplers and plot the joint posterior of $\mu$ and $\phi$. What are your comments on the results?

